---
title: 'Anonymous Functions, Part II: gsubfn'
description: "gsubfn's as.function and fn$* notation"
author: Edward Visel
date: '2018-04-07'
slug: anonymous-functions-2
type: post
categories:
  - R
  - programming
tags:
  - functions
  - r
  - package
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

_This is a follow-up to_ [Anonymous Functions, Not Variables](blog/anonymous-functions). 
_For context, read that first._

After my previous post, Brodie Gaslam pointed me in an interesting direction on 
Twitter:

```{r tweet, echo=FALSE}
blogdown::shortcode('tweet', '978250706707779585')
```

He's right. [Gabor Grothendieck's gsubfn 
package](https://cran.r-project.org/web/packages/gsubfn/index.html) is centered 
around its `gsubfn` function, an extended version of `gsub` whose `replacement` 
parameter can accept functions, e.g. to do arithmetic with numbers in a string. 

That anonymous function can be written as a formula, a capability enabled by a 
`as.function.formula` function which is also exported. It provides very similar 
functionality to the `lambda` function I defined, so the nested `map` example 
could be written

```{r}
library(purrr)

map(c("a", "b", "c"), 
    gsubfn::as.function.formula(x ~ map_chr(1:3, 
                            gsubfn::as.function.formula(y ~ paste0(x, y))))) %>% 
    str()
```

As it appears, it is a method for the S3 generic `base::as.function`, so when 
gsubfn is loaded, the syntax can be condensed to

```{r}
library(gsubfn)

map(c("a", "b", "c"), 
    as.function(x ~ map_chr(1:3, 
                            as.function(y ~ paste0(x, y))))) %>% 
    str()
```

and the generic will dispatch to it when passed a formula.

The best part of `gsubfn::as.function.formula`, though, is the alternative way 
it can be called. At the end of the previous post, I concluded 

> Ideally, it would be nice to drop the `lambda` call altogether, but as far as 
> I have been able to find, the only way to do so would be to redefine 
> `purrr:::as_mapper.default` to call `lambda` instead of `rlang::as_closure` 
> when passed a formula with anything on the left-hand side.
>
> For now, then, extra keystrokes abide.

But `gsubfn` has found a clever way to minimize keystrokes. It defines an `fn` 
object with its own class. That class has a method for `$` that is rather 
different than its usual subsetting duties. Instead, it takes a call with a 
formula in it, and returns that call with the formula converted to a function 
by `as.function.formula` (with some safeguards to ignore actual formulas). So 
while it looks a little weird, at first, you can write

```{r}
fn$sapply(1:3, 
          i ~ paste0(letters[i], i))

fn$map2_chr(letters[1:3], 
            1:3, 
            l + n ~ paste0(l, n))
```

or for the nested `map` example, 

```{r}
fn$map(c("a", "b", "c"), 
       x ~ fn$map_chr(1:3, 
                      y ~ paste0(x, y))) %>% 
    str()
```

effectively modifying functions on the fly to accept full formula lambda 
notation in a beautiful bit of code.
