---
title: Coalescing joins in dplyr
author: Edward Visel
date: '2018-07-28'
slug: coalescing-joins
type: post
categories:
  - R
tags:
  - dplyr
  - joins
  - munging
  - missing-data
description: 'Filling in missing data by joining'
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: 'Coalescing joins in dplyr'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


When aggregating data, it is not uncommon to need to combine datasets 
containing identical non-key variables in varying states of completeness. There 
are various ways to accomplish this task. One possibility an coalescing join, a 
join in which missing values in `x` are filled with matching values from `y`. 
Such behavior does not exist in current dplyr joins, [though it has been 
discussed, and so may someday](https://github.com/tidyverse/tidyr/issues/183). 
For now, let's build an `coalesce_join` function.

First, some sample data:

```{r sample-data, message=FALSE}
library(dplyr)
set.seed(47)

df1 <- data_frame(
    key  = c('a', 'b', 'c', 'd', 'e', 'f'),
    var1 = c(  1,   2,   3,   4,  NA,  NA),
    var2 = c( NA,  NA,  NA,  NA,   5,   6),
    var3 = c(  1,   2,   3,   4,   5,   6)
)

df2 <- data_frame(
    key  = c('c', 'd', 'e', 'f'),
    var1 = c( NA,  NA,   5,   6),
    var2 = c( NA,   4,   5,  NA),
    var4 = c(  3,   4,   5,   6)
)

df1

df2
```

The goal, then, is to get to:
```r
#> # A tibble: 6 x 4
#>   key    var1  var2  var3  var4
#>   <chr> <dbl> <dbl> <dbl> <dbl>
#> 1 a         1    NA     1    NA
#> 2 b         2    NA     2    NA
#> 3 c         3    NA     3     3
#> 4 d         4     4     4     4
#> 5 e         5     5     5     5
#> 6 f         6     6     6     6
```

This is still a relatively simple case, which has a single, complete key 
variable and in which the data is identical between variables, i.e. decisions 
about which data takes precedence are inconsequential. Those more complicated 
cases can be handled by an extension of the logic necessary here, though.

The core operation of `coalesce_join` will be done by `dplyr::coalesce`, which 
replaces `NA` values in a vector with corresponding non-missing values from 
another of identical length (or length 1), e.g.
```{r coalesce}
coalesce(
    c( 1, 2, NA, NA), 
    c(NA, 2,  3, NA)
)
```

The logic we want, then, is:
```{r logic}
full_join(df1, df2, by = 'key') %>% 
    mutate(
        var1 = coalesce(var1.x, var1.y), 
        var2 = coalesce(var2.x, var2.y)
    ) %>% 
    select(key, var1, var2, var3, var4)
```
...but without all the explicit references to specific column names. Thus, 
written for robustness, 

```{r coalesce_join}
coalesce_join <- function(x, y, 
                          by = NULL, suffix = c(".x", ".y"), 
                          join = dplyr::full_join, ...) {
    joined <- join(x, y, by = by, suffix = suffix, ...)
    # names of desired output
    cols <- union(names(x), names(y))
    
    to_coalesce <- names(joined)[!names(joined) %in% cols]
    suffix_used <- suffix[ifelse(endsWith(to_coalesce, suffix[1]), 1, 2)]
    # remove suffixes and deduplicate
    to_coalesce <- unique(substr(
        to_coalesce, 
        1, 
        nchar(to_coalesce) - nchar(suffix_used)
    ))
    
    coalesced <- purrr::map_dfc(to_coalesce, ~dplyr::coalesce(
        joined[[paste0(.x, suffix[1])]], 
        joined[[paste0(.x, suffix[2])]]
    ))
    names(coalesced) <- to_coalesce
    
    dplyr::bind_cols(joined, coalesced)[cols]
}
```

Let's try it out:

```{r test}
coalesce_join(df1, df2, by = 'key')
```

Beautiful!

While this code could be used for an updating join, because of its structure, 
it will always prioritize the data from `x` over `y`. More flexible (updating 
columns from specified sources) or strict (checking equality of non-missing 
data in coalesced columns) versions could be written, but the former would 
require a good API for specifying data precedence, and the latter would be 
inconsistent with the behavior of `coalesce` itself.

Adapt as you like.
