---
title: Anonymous Functions, Not Variables
description: "Use the LHS of a formula to specify variable names in purrr-style lambda functions"
author: Edward Visel
date: '2018-03-25'
slug: anonymous-functions
type: post
categories:
  - R
  - programming
tags:
  - functions
  - purrr
  - rlang
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
knitr::knit_engines$set(racket = function(options){
    engine = options$engine
    code = paste("-e", shQuote(paste(options$code, collapse = "\n")))
    code = paste(options$engine.opts, code)
    cmd = system('which racket', intern = TRUE)
    out = if (options$eval) {
        message("running: ", cmd, " ", code)
        tryCatch(system2(cmd, code, stdout = TRUE, stderr = TRUE, 
            env = options$engine.env), error = function(e) {
            if (!options$error) 
                stop(e)
            paste("Error in running command", cmd)
        })
    } else ""
    if (!options$error && !is.null(attr(out, "status"))) 
        stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, options$code, out)
})
```

I am a very heavy [purrr](https://github.com/tidyverse/purrr) user. The killer 
feature is clearly `map_df` (fairly recently rebranded as `map_dfr` and 
`map_dfc` for row and column binding, respectively) to iterate over a list à la 
`lapply` and simplify the result to a data frame. Thanks to the power of 
`dplyr::bind_rows`, it fixes all the drawbacks of `sapply`'s `simplify2array` 
behavior:

* It returns a data frame, not a matrix or array, so multiple types can be 
  kept. Never see a matrix of lists again.
* Variables are aligned by names, not locations, and thus will not be recycled 
  to locations they do not belong, and `NA`s will be inserted as appropriate.
* Turning the list names into a column is as simple as passing a name for it to 
  the `.id` parameter.
  
Along with tidyr's `gather` and `spread`, `map_df` is indispensable for me—to 
do without would mean a lot of `do.call(rbind, ...)` that can go very wrong 
without warning.

purrr has lots of other lovely functions, too, like `transpose`, `partial`, and 
`cross` that are well worth learning. At the core of its functional programming 
abilities, though, is `as_mapper`, a function unlikely to be called much 
directly, but which is called by every `map` variant to decide what to do with 
the `.f` and `...` parameters, which may either be parameters for subsetting 
each element or a function, possibly with additional parameters. In the latter 
case, `purrr:::as_mapper.default` is [now] simply a wrapper for 
`rlang::as_closure`.

This function enables one of purrr's far-reaching conveniences: the ability to 
define anonymous functions with formulas. In its most basic format, e.g. a 
unary function for `map`, `function(x) as.character(x)` can be rewritten as 
`~as.character(.x)`. The `~` both quotes the function, and can be read for the 
unary case as `function(.x)`. 

For unary functions, the input can be referred to as `.x`, just `.` (which I
generally avoid to eliminate confusion with data piped in), or `..1`, since it
is the first parameter. For binary functions like those for `map2` or `reduce`,
the inputs are typically referred to as `.x` and `.y` (though the first behaves
the same as that of a unary function, and the second, as such, can also be
referred to as `..2`). For a polyadic function like that for `pmap`, the
parameters are input as `...`, which can be collected, e.g. `params =
list(...)`, or accessed directly by number with `..1`, `..2`, etc. notation as
usual.

While this notation has a small learning curve, the convenience when composing 
anonymous functions adds up. Keystrokes are saved. Less noticeably but perhaps 
more importantly, the mental effort of naming parameters is eliminated. ("Did I 
call something `x` already?")

This is all very good. However, there are limitations. In particular, nested 
calls with this notation will typically not work, because the variable is called 
the same thing, e.g.

```{r}
library(purrr)

map(c("a", "b", "c"), 
    ~map_chr(1:3, 
         ~paste0(.x, .x))) %>% 
    str()
```
Assuming such a case where double iteration is necessary and cannot be
vectorized away^[Here, `outer(c("a", "b", "c"), 1:3, paste0)`, though that does
return a matrix instead of a list.], there is simply no convenient way to write
this code with this notation. If the function is supposed to be passed to the
first parameter of one of the functions, an alternative is to pass the function
and other parameters separately, e.g.

```{r}
map(c("a", "b", "c"), 
    ~map_chr(1:3, 
             paste0, 
             .x)) %>% 
    str()
```
but this gets hard to read, and the order cannot be reversed without flipping 
the loops, which may be impractical or impossible. Thus, the alternative is to 
fall back to standard anonymous function notation so as to specify different 
names for each parameter:

```{r}
map(c("a", "b", "c"), 
    function(x) {
        map_chr(1:3, function(y) paste0(x, y))
    }) %>% 
    str()
```
This works, but loses a lot of convenience for the sake of avoiding a name clash. 
But I think there is a better way.

The functions in question are [anonymous
functions](https://en.wikipedia.org/wiki/Anonymous_function), i.e. they are not
assigned to a name, but are rather raw expressions that define a function. In R,
the most common usage of such functions (sometimes called _lambda expressions_
due to their origins in Alonzo Church's [lambda
calculus](https://en.wikipedia.org/wiki/Lambda_calculus)) is when passing a
function as a parameter to a function like `map`. Since their introduction in
Lisp, anonymous functions have become a part of most modern programming
languages. Syntax differs, of course; for example in Racket, a modern Lisp:

```{racket}
((lambda (x) 
  (* x x)) 
 2)
```

or Python:
```{python}
(lambda x: print(x * x))(2)
```

or Julia:
```{julia}
((x) -> x * x)(2)
```

or R:
```{r}
(function(x) x * x)(2)
```

Comparing these syntaxes to that of purrr's anonymous functions, a disadvantage
of the latter becomes apparent: there is no way to specify parameter names,
which is why nested evaluation is impossible. That said, given both a notation
like that of Julia above and how models in R use formula notation, it seems
plausible that variable names could be specified on the otherwise unused
left-hand side (LHS) of a formula function.^[The fascinating magrittr
alternative [pipeR](https://github.com/renkun-ken/pipeR) implements [something
of the sort for naming data piped into a 
function](https://renkun-ken.github.io/pipeR-tutorial/Pipe-operator/Pipe-by-formula.html).]
Multiple variables cannot be separated by commas (which would terminate the
formula), but could be separated by `+`, like multiple terms in a regression.

To build such capability, let's use [rlang](https://github.com/tidyverse/rlang), 
which has handy utilities for parsing and creating formulas and functions. 

* The variable names can be parsed from the LHS (returned as an expression by 
  `f_lhs`) with `base::all.vars`, which takes an expression and returns a 
  character vector of variables it uses.
* The function can be created with `new_function`, which requires
  * the parameters, passed as a named list. Since there should be no defaults, 
    the `args` parameter should be a list with named empty elements. For known 
    inputs, such a list can be created with `alist`, e.g. `alist(x = , y = )`, 
    but because of its quoting semantics, it is very difficult to program with. 
    A suitable list can be generated by simply making a list of empty 
    expressions, though.
  * the body, which is simply the right-hand side of the formula, extracted 
    with `f_rhs`, and 
  * the environment. Formulas, like functions, have environments, which can be 
    accessed with `f_env`.
    
All together,

```{r lambda, message=FALSE}
library(rlang)

lambda <- function(x){
    arg_names <- all.vars(f_lhs(x))
    new_function(args = set_names(rep(list(expr()), 
                                      length(arg_names)), 
                                  arg_names), 
                 body = f_rhs(x), 
                 env = f_env(x))
}
```

Now we can write formula functions in a fashion fully equivalent to ordinary 
ones, e.g.
```{r}
lambda(x ~ x * x)

lambda(x ~ x * x)(2)
```

It works for multiple parameters, too:
```{r}
lambda(x + na.rm ~ mean(x, na.rm = na.rm))(c(1, NA, 3), na.rm = TRUE)
```

Now, in the original nested `map` example, 
```{r}
map(c("a", "b", "c"), 
    lambda(x ~ map_chr(1:3, 
                       lambda(y ~ paste0(x, y))))) %>% 
    str()
```

Ideally, it would be nice to drop the `lambda` call altogether, but as far as I 
have been able to find, the only way to do so would be to redefine 
`purrr:::as_mapper.default` to call `lambda` instead of `rlang::as_closure` 
when passed a formula with anything on the left-hand side.

For now, then, extra keystrokes abide.


