---
title: Fireworks
author: Edward Visel
date: '2018-11-11'
slug: fireworks
type: post
categories:
  - R
  - dataviz
  - animation
  - generative art
tags:
  - r
  - ggplot2
  - gganimate
  - generative art
description: 'Animating fireworks with ggplot2 and gganimate'
featured: 'firework-banner.png'
featuredalt: 'firework'
featuredpath: 'date'
linktitle: 'Fireworks'
---

Since [Thomas Lin Pedersen](https://twitter.com/thomasp85) took over 
[gganimate](https://github.com/thomasp85/gganimate), I've been building 
animations. Mostly what I've built is not for any particular data visualization 
purpose. My motivations vary, but have included 

- I want to try out new features of the package
- I'm bored
- I'm trying to understand trig functions in polar space
- I have an idea
- I can't focus
- I want to play with matrix transformations
- I want to make pretty things

...but I mostly make them because they make me happy. The results could be 
charitably called generative art, though they're more a pale homage to artists 
deserving of the title like 
[\@beesandbombs](https://twitter.com/beesandbombs).

Today I went back to an old idea stemming from some work with spatial grids in 
[sf](https://r-spatial.github.io/sf/). Plotting the grids with a projection, I 
discovered that when covering the whole globe, projected grids sometimes look 
like fireworks, e.g.

```{r sf-grid, message=FALSE}
library(tidyverse)
library(sf)
theme_set(hrbrthemes::theme_ipsum_ps())

grid_plot <- expand.grid(lon = seq(-180, 180, 10), 
                         lat = seq(-90, 90, 10)) %>% 
    pmap(~st_point(c(...))) %>%    # convert each pair to an sf point
    st_sfc(crs = 4326) %>%    # convert all points to sf column
    st_sf() %>%    # convert to sf data frame
    ggplot() + 
    geom_sf()

grid_plot

grid_plot + coord_sf(crs = "+proj=laea +lat_0=-90 +ellps=WGS84 +no_defs")
```

Grids as fireworks sounded like a prime candidate for animation, which I 
finally got back to today. Instead of using sf, I stuck to polar coordinates, 
which can produce a similar effect with less effort.

Most of my plots were not what I wanted. Some didn't move. Some moved too slow, 
or with the wrong acceleration. Some were just expanding rings. But here I 
present some of my more interesting failures, together with a couple results 
that could be convincingly called fireworks.

***

## A hole

Exponential decay doesn't slow down, it goes backwards:
```{r gganimate, warning=FALSE}
library(gganimate)
theme_set(theme_void() + theme(
    panel.background = element_rect(fill = 'black')
))
```
```{r hole, cache=TRUE}
p <- crossing(x = 1:30, nesting(t = 1:10, y = .5^(seq(t)))) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.5)

animate(p, fps = 30)
```

## Straight

```{r straight, cache=TRUE}
p <- map_dfr(1:10, ~crossing(x = 1:30, nesting(
        y = seq(1, .x, length.out = 10)^0.5, 
        t = 1:10
    ))) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.3)

animate(p, fps = 30)
```

I thought this was what I wanted, but seeing it, while it may be how we think 
fireworks look (a bunch of points shot out at equally-spaced intervals on a 
circle), it doesn't actually look right, because fireworks aren't circles, 
they're spheres.

## Particles and gnats

I couldn't figure out the spacing, so I tried random uniform numbers instead of 
a sequence:

```{r particles, cache=TRUE}
p <- map_dfr(1:10, ~data_frame(y = seq(1, .x, length.out = 10), t = 1:10)) %>% 
    mutate(x = runif(n())) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.5)

animate(p, nframes = 300, fps = 30)
```

...which returned a bunch of weird particles, or with shorter timing

```{r gnats, cache=TRUE}
animate(p, nframes = 30)
```

..gnat-swarm fireworks. Which are fireworks, but not the type I wanted.

## Mandala

I don't know what happened here, but it's trippy.

```{r mandala, cache=TRUE}
p <- map_dfr(1:10, ~crossing(
        x = 1:30, 
        nesting(
            y = runif(10), 
            t = 1:10
        )
    )) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.5)

animate(p, nframes = 20)
```

## Explosion

Ok, this is what I meant, but it just looks like an explosion.

```{r explosion, cache=TRUE}
p <- map_dfr(1:10, ~crossing(
        x = runif(30), 
        nesting(
            y = seq(1, .x, length.out = 10)^0.5, 
            t = 1:10)
        )
    ) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.5)

animate(p, fps = 30)
```

## Spirals

Getting pretty close.

```{r spirals, cache=TRUE}
p <- map_dfr(1:10, ~crossing(
        x = {
            x = seq(30) + 0.3*.x; 
            ifelse(x > 30, x - 30, x)
        }, 
        nesting(
            y = seq(1, .x, length.out = 10)^0.5, 
            t = 1:10)
        )
    ) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.3)

animate(p, fps = 30)
```

## In-n-out

Wait what happened here
```{r in-n-out, cache=TRUE}
p <- map_dfr(0:10, ~crossing(
        x = {
            x = seq(30) + 0.6*.x; 
            ifelse(x >= 30, x - 30, x)
        }, 
        nesting(
            y = seq(1, .x, length.out = 10)^0.5, 
            t = 1:10)
        )
    ) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.5)

animate(p, fps = 30)
```

## Firework

```{r firework, cache=TRUE}
p <- map_dfr(1:10, ~crossing(
        x = {
            x = seq(30) + 0.6*.x; 
            ifelse(x > 30, x - 30, x)
        }, 
        nesting(
            y = seq(1, .x, length.out = 10)^0.5, 
            t = 1:10)
        )
    ) %>% 
    ggplot(aes(x, y)) + 
    geom_point(color = 'white') + 
    coord_polar() + 
    transition_time(t) + 
    shadow_wake(0.3)

animate(p, fps = 30)
```

Yep, that's definitely a firework. Done.
