---
title: match.arg
author: Edward Visel
date: '2019-07-04'
slug: match.arg
categories:
  - R
  - programming
tags:
  - functions
  - utility
description: 'A weird function that can help you write better functions'
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: 'match.arg: A weird function that can help you write better functions'
type: post
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


Lately I've been working with a lot of people whose first language is not R, 
which has given me more of an appreciation for R's oddities. Some in retrospect 
were probably ill-advised, like partial matching with `$`:

```{r partial-matching}
mtcars$disp
mtcars$di
mtcars$d    # Bonus points if you knew what this would do.
```

Some are comparatively weird, but function fine, like that R doesn't care if 
keyword arguments come before positional ones; it just extracts keywords first 
and then matches positions:

```{r positional-args}
mean(x = c(1, NA, 3), 0, TRUE)
mean(na.rm = TRUE, c(1, NA, 3))
```

But some are weird, but actually really convenient and great, which brings us 
to the subject of this post: `match.arg`.

`match.arg` is a function that only works in functions, and thus is not itself 
a function people tend to encounter until they try to write code for other 
people. (You _could_ use it in functions you write for yourself, but it matters 
less, for reasons that will become apparent in a moment.) It's also a function 
that's baked into much of R's DNA, so to speak.

For instance, if you look at the documentation for `?optim`, base R's 
general-purpose optimization function, its "Usage" section looks like this:

```r
optim(par, fn, gr = NULL, ...,
      method = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN",
                 "Brent"),
      lower = -Inf, upper = Inf,
      control = list(), hessian = FALSE)
```

Something there is a little funny: Why is the default of `method` a vector of 
six methods? Is it going to run six different ways? The argument documentation 
seems to suggest otherwise:

> `method`    The method to be used. See ‘Details’. Can be abbreviated.

"Details" confirms:

> The default method is an implementation of that of Nelder and Mead (1965), 
> that uses only function values and is robust but relatively slow. It will 
> work reasonably well for non-differentiable functions.

It then describes `"BFGS"`, `"CG"`, etc. as alternative methods, which makes 
sense—the documentation is telling us all the possible alternatives. But 
there's still something weird here: Is the default really a vector of length 6? 
Or is that just a documentation nicety, and the signature of the function 
_actually_ contains `method = "Nelder-Mead"`?

A quick investigation reveals that it's not just a nicety—the actual default is 
a vector of length 6

```{r optim-args}
args(optim)
```

...but somehow that turns into a default of `"Nelder-Mead"`. How? The culprit 
is a few lines into the body of the function:

```{r optim-body}
head(optim, 8)
```

What does 
```r
method <- match.arg(method)
```
do?

Starting at `?match.arg`, it tells us it does "argument verification using 
partial matching". More particularly, 

> `match.arg` matches `arg` against a table of candidate values as specified by 
> `choices`, where `NULL` means to take the first one.^[This `NULL` bit is 
confusing, even if you come back after reading everything. It refers to a call 
like `match.arg(NULL, c("deviled eggs", "potato salad"))`, but the function 
rarely gets used that way. The "take the first one" bit matters, though, 
because this is the behavior if `choices` isn't passed, as is explained later.]

The "Usage" section confirms that `optim` _definitely_ isn't running six 
methods by default:

```r
match.arg(arg, choices, several.ok = FALSE)
```

Ok, this seems to make some sense—`match.arg` checks whether an argument to a 
function is within a list of possibilities. Let's try it out:

```{r check-args, error=TRUE}
check_is_us_flag_color <- function(color){
    match.arg(color, c("red", "white", "blue"))
}

check_is_us_flag_color("blue")
tryCatch(
    check_is_us_flag_color("orange"),
    error = identity
)
```

Ooh, look, we even get informative error messages! How nice.

It's supposed to do partial matching too. Let's try: 

```{r}
check_is_fourth_of_july_activity <- function(activity){
    match.arg(activity, choices = c(
        "watch a parade",
        "barbeque in the park",
        "watch fireworks"
    ))
}

check_is_fourth_of_july_activity("watch fireworks")    # Still works.
check_is_fourth_of_july_activity("barbeque")    # Works too!
tryCatch(
    check_is_fourth_of_july_activity("watch"),    # Doesn't work. That's good.
    error = identity
)
```

This could be useful! It encourages us to use descriptive options like 
`"Nelder-Mead"`, but if people don't want to type that all the time, they can 
just type the unambiguous `"Nelder"`, and that will work fine. Also, such a 
sophisticated approach still takes very little effort from the person writing 
the function, which is awfully nice.

<details><summary>
If you want to understand the details of precisely _how_ `match.arg` handles 
partial matches</summary>, the documentation says

> Matching is done using `pmatch`, so `arg` may be abbreviated.

Clicking through and trying out `pmatch` shows it's pretty simple—it's not 
fuzzy matching _à la_ `agrep`, just matching from the beginning of strings.
</details>

Ok, now we've got a grasp of how `match.arg` works, but that still doesn't 
explain how that line from `optim` works:

```r
method <- match.arg(method)
```

There are no choices specified! `?match.arg` gives us a hint:

> In the one-argument form `match.arg(arg)`, the choices are obtained from a 
> default setting for the formal argument `arg` of the function from which 
> `match.arg` was called. (Since default argument matching will set `arg` to 
> `choices`, this is allowed as an exception to the ‘length one unless 
> `several.ok` is `TRUE`’ rule, and returns the first element.)

There's a lot going on here. This directly answers one question: What happens 
if `method` is not set? In this case, it will return the first element, which 
is why the default is in fact `"Nelder-Mead"`. 

_Even if you never use `match.arg`, this behavior is a good thing to 
understand,_ because you'll see documentation like this everywhere: 
`?read.table`, `?png`, `?order`, `?t.test`, `?ggplot2::position_dodge`, 
`?tidyr::fill`, `?data.table::shift`, etc., etc., etc.

But there are two possibilities here:

- no argument is passed, and `match.arg` gets the full length-6 vector, or
- an argument is passed, and `match.arg` is only passed one string.

Taking the first element of a vector explains how the first possibility works:

```{r first-is-default}
check_good_firework <- function(firework = c('bottle rocket', 'roman candle', 'ones that require a license')){
    match.arg(firework)
}

check_good_firework()
```

`match.arg` is passed the full vector for `firework`, and picks the first one. 
But is it using that vector for choices? Let's see:

```{r first-is-default-2}
check_good_firework(c('bottle rocket', 'roman candle', 'ones that require a license'))
tryCatch(
    check_good_firework(c('snake', 'M80')), 
    error = identity
)
```

Hmm, so it _is_ getting the choices from somewhere. That's good, as it explains 
how the case in which an argument is passed is handled. But where is 
`match.arg` getting the options if they aren't passed to it? The docs, one more 
time:

> In the one-argument form `match.arg(arg)`, the choices are obtained from a 
> default setting for the formal argument `arg` of the function from which 
> `match.arg` was called.

Ah, "the formal argument `arg` of the function from which `match.arg` was 
called". But what's a "formal argument"? The details get a little 
hairy^[Technically only closures have formals, not primitive functions. Go read 
about the difference if you like, but it rarely matters from a user 
perspective.], but essentially formals are the stuff you put in the parentheses 
after `function`. There's even a sensibly-named function to get them, should 
you need:

```{r formals}
str(formals(optim))
```

Notice this is not quite a normal list—some elements are empty (like `par` and 
`fn`), and `...` is actually the name of an element (that doesn't exist). It is 
a list, but this is the topsy-turvy world of operating on the language, so this 
is a _pairlist_, which only much ever gets used for messing with the structure 
of functions.^[Try playing around with `alist` and you'll start to see what's 
going on.]

Regardless, we can see that `method` is there with its default. The source of 
`match.arg` shows how it gets the default to use for `choices`:

```{r}
head(match.arg, 7)
```

This code is both complicated and hairy (please don't use `<-` in function 
calls), but we can see that `sys.function` gets the calling function, and 
`formals` extracts the formal arguments. Lines 5-6 extract the argument from 
those formals.

Ignoring the environment-handling stuff, let's try it out:

```{r sys.function}
show_function <- function(x, y = c("foo", "bar")){
    sys.function()
}

show_function()

fmls <- formals(show_function())
str(fmls)

fmls$y    # This is still a language object...
eval(fmls$y)    # ...so evaluate it to make it a real object
```

Cool! This code explains the remaining mystery of `match.arg`: When only passed 
an argument, it grabs the default argument for that parameter and uses it as 
`choices`. That also explains why `match.arg` only works in a function if 
`choices` is not specified:

```{r not-in-function}
tryCatch(
    match.arg(c("bratwurst", "hot dog", "hamburger")),
    error = identity
)
```

So when should you use `match.arg`? When you're writing a function with a 
parameter that can take one of a fairly small number of string values. In 
return, 

- your function will error informatively if passed an incorrect value (instead 
of breaking wherever it gets used),
- users get convenient partial-matching,  
- if you put the choices as the default, the possible choices will be very 
clear (to people who know about `match.arg`), and
- your friends coming from Python will reinforce their belief that R is 
strange, but will be unable to reproduce the behavior in Python without 
significant code.

So is R quirky? Yep. But that's not all bad.
