---
title: 'Mapping leaves'
author: Edward Visel
date: '2019-01-20'
slug: recursion
categories:
  - R
  - programming
tags:
  - r
  - functional-programming
  - purrr
  - utility
description: 'Recursive functional programming'
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: 'Mapping leaves: Recursive functional programming'
type: post
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

## `at_depth`

I love [purrr](https://purrr.tidyverse.org/).^[I'm also very fond of `Map` and
`Reduce` and so on, so maybe I just like functional programming.] Aside from 
[its anonymous function 
notation](https://alistaire.rbind.io/blog/anonymous-functions/), one of the
functions that made me love the package was `at_depth`, which iterates across a
list at a specified level of nesting. It has since been deprecated in favor of
`modify_depth`, which is more powerful, but is significantly more finicky.

The additional power is because the `.depth` parameter can now be passed a
negative integer to index up from the bottom of the list. When I first heard
this, I was excited, because it sounded like it could iterate across the ragged
leaf nodes of a list. After months of trying, I still could not get it to work.
I discovered that was because it does not iterate across leaf nodes, but instead
selects depth by indexing up from the deepest level of the list, iterating over
whatever nodes are at that level regardless of whether they are lists or not.

## `rapply`

Part of my initial excitement upon hearing that `modify_depth` could take
negative `.depth` values was because I thought it may be less-confusing version
of `rapply`, the second-most confusing function in base R after `reshape`.^[I
can now get `rapply` to work about half the time, and even used it usefully
once. I have never gotten `reshape` to do what I want. I've seen other people
make it work, so I know it *can* function, but for the life of me I can't
understand its parameters.] `rapply` is a recursive version of `lapply` which
(in theory) lets you iterate a function across the leaf nodes of a list.

For example:

```{r rapply}
library(magrittr)

l <- list(1, list(2:3, 4, list(5:6), 7))

rapply(l, function(x) x + 1, how = "replace") %>% 
    str()
```

With the default `how = "unlist"`, `unlist` is called on the result, collapsing it:

```{r rapply-unlist}
rapply(l, function(x) x + 2)
```

The biggest problem with `rapply` is that it decides whether to recurse over a node or apply the function based on the behavior of `typeof` and `is.list`, not those of `class`. Consequently, it cannot operate on objects that are lists, like data frames or models—it will recurse right into them:

```{r rapply-data-frame}
rapply(list(list(mtcars)), mean)
```

## Mapping over leaf nodes

So let's build our own recursive map/apply function that can manage list
objects. Sticking to purrr's structure, we'll make the function an S3 generic
with a default method that will do the recursion. Objects can be differentiated
from unclassed lists with `rlang::is_bare_list`, and `purrr::as_mapper` will let
it handle the same variety of input structures as `map` and friends.

```{r map_leaves}
map_leaves <- function(.x, .f, ...){
    UseMethod("map_leaves")
}

map_leaves.default <- function(.x, .f, ...){
    if (rlang::is_bare_list(.x)) {
        purrr::map(.x, map_leaves, .f, ...)    # recurse!
    } else {
        .f <- purrr::as_mapper(.f, ...)
        .f(.x, ...)    # call function on node
    }
}
```

Let's try it out:

```{r map_leaves-demo-1}
l %>% map_leaves(~.x + 1) %>% str()

ltcars <- map_leaves(l, ~mtcars[.x, ])
ltcars
```

Both of these could be done with `rapply`. But `rapply` can't operate upon the 
data frames of `ltcars`, whereas `map_leaves` can:

```{r map_leaves-demo-2}
map_leaves(ltcars, dplyr::summarise_all, mean)
```

If we want to do the same thing in pure base R, it looks similar:

```{r}
leafapply <- function(X, FUN, ...){
    if (is.list(X) && is.null(attr(X, "class"))) {
        lapply(X, leafapply, FUN, ...)
    } else {
        FUN(X, ...)
    }
}

l %>% leafapply(`+`, 1) %>% str()

ltcars <- leafapply(l, function(x) mtcars[x, ])
ltcars

leafapply(ltcars, function(x) aggregate(. ~ 1, x, mean))
```

## Collapsing lists but not list objects

Neither `map_leaves` nor `leafapply` can condense the results like `rapply` 
does with `how = "unlist"`, though. `unlist` itself won't work, because it 
also ignores list objects. To collapse recursively, we need to write a smarter 
version of `unlist`.

In this case, we need to recurse over sublists of the input (and, because it is
recursive, potentially sublists of sublists), and then collapse the recursively
flattened list. We can use a similar test to `map_leaves` and `leafapply` to
identify sublists, but we also need a function to collapse each flat list. To
imitate `unlist`, this would be `c`, applied with `purrr::invoke` or `do.call`.
But it makes sense to make this function user-settable, because for other data
types, different collapsing functions will allow more forms of collapsing.

```{r deflate}
deflate <- function(.x, .f, ...){
    UseMethod("deflate")
}

deflate.default <- function(.x, .f, ...){
    .f <- purrr::as_mapper(.f)
    is_sublist <- purrr::map_lgl(.x, rlang::is_bare_list)
    .x[is_sublist] <- purrr::map(.x[is_sublist], deflate, .f, ...)
    purrr::invoke(.f, .x, ...)
}
```

`deflate` can imitate `unlist`:

```{r deflate-c}
deflate(l, c)
```

But because it can take different collapsing functions, it can also do fancier  
collapsing more appropriate for, say, data frames:

```{r deflate-bind_rows}
deflate(ltcars, dplyr::bind_rows)
```

Again, we can write the same thing in base R:

```{r Deflate}
Deflate <- function(f, x, ...){
    is_sublist <- vapply(x, function(y) is.list(y) && is.null(attr(y, "class")), logical(1))
    x[is_sublist] <- lapply(x[is_sublist], Deflate, f = f, ...)
    do.call(f, c(x, ...))
}

Deflate(c, l)
Deflate(rbind, ltcars)
```

Exciting!

## Next steps

Some possible extensions:

- Both functions could be combined into one to mimic `rapply` more directly. I
like the modular, composable form, though; it fits [the Unix
philosophy](https://en.wikipedia.org/wiki/Unix_philosophy) nicely.
- A `class` parameter *à la* `rapply` could be added to `map_leaves` (and made 
way more useful than `rapply`'s). Presumably for `map_leaves`, nodes that don't 
match the class should be ignored; for `deflate` they should be subsetted out.
- Make a version of `eapply` that takes a `class` parameter.
- A version of `map_leaves` that indexes up from each leaf node could be 
created. While powerful, this has a lot of potential to be finicky to use.
- Make new methods for similar objects like dendrograms *à la* `dendrapply`.
- Put all these functions in a package, if people will use it.
- Something else? Add a comment or ping me [on Twitter](https://twitter.com/alistaire)!
